---
title: "Introducing R and RStudio"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This page contains information on how to install R and RStudio, what it is used for and some basic termonogoly and commands, like installing packages.

## Explanation

### What is R?
R is a software environment in which you can analyze data by typing instructions (programming). In the field of behavioral and social sciences, researchers and teachers are increasingly working with R instead of programs like STATA, SAS, SPSS and EXCEL, because R has some advantages compared to these programs:

* it is free, so you can also use it after you finished your studies;
* it is open-source and researchers all over the world write code that is transparent and which can beused for free;
* data analysis and data visualization possibilities in R are almost endless;
* the language is extremely flexible.

R allows you to manage data sets, to change data in order to facilitate analysis, to analyze data and to visualize data. R also allows you to integrate data in text files, which reduces the amount of ‘copy-pasting’ and reduces the number of mistakes made in this process, thus improving the reproducibility of research.

### What is RStudio?

<iframe src="https://www.youtube-nocookie.com/embed/riONFzJdXcs?modestbranding=1&rel=0" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" sandbox="allow-scripts allow-same-origin" allowfullscreen="" width="500" height="281"></iframe>

RStudio is a ‘shell’ designed to help you to be more productive with R. RStudio is an Integrated Development Environment (IDE) that helps you develop programs (and scripts) in R. RStudio has a set of integrated toolsthat are helpful when analyzing and visualizing your data. More information about RStudio can be found here: https://rstudio.com/products/rstudio/features/

### Why R and RStudio at BMS?
At BMS we think students in the behavioral and social sciences should have a clear understanding of ‘data’ and using R and RStudio will give you both a clear understanding of data and a flexible and transparent way to handle these data. R also allows us to teach statistics in a flexible way, adding new ideas (like ‘big data analysis’) to our curriculum more easily.

### Installing R and RStudio on your laptop

<iframe src="https://www.youtube-nocookie.com/embed/cX532N_XLIs?modestbranding=1&rel=0" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" sandbox="allow-scripts allow-same-origin" allowfullscreen="" width="500" height="281"></iframe>

For most purposes, installing R and RStudio on your laptop is straightforward. For R go to

* https://cran.uni-muenster.de

For R studio go to:

* https://www.rstudio.com/products/rstudio/download/#download

and select the free version.

### How to learn R?
Although we will offer you a lot of materials to learn the R basics, keep in mind that for learning R even better you need to find your way on the web to learn additional procedures and to practice even more. How?

Some suggestions:

* use a search engine (like Google) to ask questions like “how do I enter data into R?”;
* use stackoverflow (https://stackoverflow.com) to check out answers to similar questions;
* find your own pet project (data with your running times, for example, or gpx files, or COVID-19 data);
* read blogs in which data scientists show what can be done with R.

A nice feature in R is the help function. If you need help on a function, just type a ? in front of a function.If you type ?? in front of the function, you will even get help from all over the web.

```{r eval=F}
?tidyverse
??tidyverse
```

### Terminology: working directory, console, script and more

<iframe src="https://www.youtube-nocookie.com/embed/UYclmg1_KLk?modestbranding=1&rel=0" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" sandbox="allow-scripts allow-same-origin" allowfullscreen="" width="500" height="281"></iframe>

Learning R includes learning a lot of new words like ‘working directory’, ‘script’ and ‘console’. In this section some of the most commonly used terms will be introduced. Although it might be hard to understand the terms completely by reading the explanations below while not yet working in R, you will see that in a few weeks you will understand exactly what is meant with the terms.

The _working directory_ on your laptop or pc is the folder where all kinds of data, output (including pictures),and scripts are stored (as files), so you can quickly find them back later. It is wise to create a working directory for each separate course or data project and to tell R you will use that directory for this specific project.

```{r eval=F}
setwd("[...]")
# The [...] has to be replaced by something applicable to your computer.

setwd("/Volumes/myname/Userdata/My Documents/Documenten/Research/2020_Local_elections")

# The current working directory (in case you forget) can be found with the command
getwd()
```

Make sure you can find the path of a working directory in your mac or pc: 

- on Mac, in Finder, the small ‘radar’ on top allows you to copy the location as ‘pathname’ 
- in Windows, in the File Explorer, select View in the toolbar, -> Options, select Change folder and search options, to open the Folder Options dialogue box, Click View to open the View tab. In Advanced settings, add a check mark for Display the full path in the title bar, Click Apply, Click OK to close the dialogue box.

A scriptis a set of commands, preferably with some _#comments_ explaining what happens in the command lines. In order to distinguish between command lines and comments, comments are preceded by the # sign.

Use - - - - - to break up your script into easily readable chunks.

If you want to run a line of the script, you move the cursor to the respective line and you use ctrl+enter (cmd+enter on a Mac). The instruction will then then automatically move to the R Console and the cursorin the script automatically jumps to the next line.

The _console_ panel (at the bottom in RStudio, with the prompt “>”) can be used to directly type commands. For example, when using R as a simple calculator you can just type in the numbers in the console.

```{r eval=F}
# adding 1 + 1
> 1+1
[1] 2
```

_Objects_, including imported data, are stored in the _global environment_. When you create new objects, these objects are stored in the global environment too. For example, if you create a frequency table, you can store the table in the global environment. It can always be called back later in the analysis. This global environment can be saved separately and called later, although for many practical purposes it is sufficient to create a global environment anew, by using the script which created that environment in the first place. The global environment can be found on the right upper side of RStudio.

Data sets, numbers, and output can be stored as objects in the global environment, using the _(left) assignment operator_ “<-”.[^1] An object can be a lot of things. For example, if you use R to create a boxplot, the boxplot can be stored as an object, to be called later when needed. When naming objects, use only lowercase letters, numbers, and \_. Use underscores (\_) to separate words within a name.

[^1]:
To be fair, also “=” will work, but most people now use the “<-” assignment operator

```{r eval=F}
# for example, to use the 10 cases throughout a script, you can define the object 'n' to '10'
n <- 10

# or to create a simple vector of numbers, you can write ...
dataset1 <- c(3, 2, 1, 4, 9, 6, 7, 100)

# calling this object in a script or in the console gives the contents of the thus created object.
dataset1
# this give the following output in the console:
```
```
[1]   3   2   1   4   9   6   7 100
```

_Functions_ in R are ‘commands’ telling R what to do: they are ‘do this’ statements. R has a lot of those. Functions end with parentheses: “()”.[^2] For example:

[^2]:
The `magrittr` package (included in the `tidyverse` package) allows you to omit parentheses () on functions that don’t have arguments. Avoid this feature, because it may be confusing. Be consistent in that functions always have parentheses, objects don’t.

```{r eval=F}
# The command head() is'Show me the first part (the head) of a dataset'.  In this example the called dataset is an object called dataset1.
head(dataset1)

# After installing the 'tidyverse' package (a package we will use at the UT, the word **package** will be explained below) and the associated pipe command ( %>%# ), use ...

dataset1 %>% head()
```

### Data frames and data types
_Data frames_ in R are matrices in which the columns are variables and the rows are units of observation (observations). Sometimes this data frame is called adata matrix, but in the R language a _data matrix_ is a data frame with only numbers, not with strings (text) or other types of variables. One single column (or: a variable) with numbers or strings is called a _vector_.[^3] Sometimes objects are not just data frames, but a data frame and some more. These objects are called _lists_.

[^3]:
Actually a vector can contain numbers, or strings, or factors, or integers, etc. One single vector just cannot have different types of data.

For reasons beyond the scope of this short introduction, some functions only work with data frames (even if they contain only numbers), not with matrices. In these cases you have to explicitly tell R your matrix is a data frame.

Variables in a data frame have a name (like ‘x’ or ‘gender’). These can be be called directly using the ‘\$’ sign. So `data$gender` refers to the column/variable gender in the data frame. More generally, the ‘\$’ sign refers to a ‘component’ of an object. If you store output as an object to the global environment, you can call a specific element of that output using the ‘\$’ sign.

There are different types of variables in a data frame. The most important are:

* a variable containing logical values (True or False) (used for some dummy variables);
* a variable containing a limited set of attributes (‘low’, ‘middle’ and ‘high’ for example) (used for nominal and ordinal variables). A factor can also be stored as an ‘ordered factor’) to change it from merely nominal to ordinal.
* a variable containing integer values (-1, 0, 1, 2, 3 etc) (mainly used for ordinal variables and for count variables);
* a variable containing real values (1.001, 1.002, 10000) (used for interval and ratio variables);
* a variable containing characters/ words (‘male’, ‘female’) (if these variables are used for dummies and for nominal variables, you have to change them to ‘factor’ variables).

<!-- Maybe it's an idea to add how these data types are called, instead of just their description. So the names you get back from the class() command. So logical, factor, numeric and character. Since these terms seem to be used later on -->

### Operators in R
The most frequently used operators beyond `-` (Minus), `+` (Plus), `*` (Multiplication), `/` (Division) and `<-` (the Left assignment operator), are:

```{r echo = F, results = 'asis'}
library(knitr)
kable(data.frame(
  operator = c("^", "!", ">", "==", ">=", "<=", ":", "$"),
  meaning = c("Exponentiation", "Not", "Greater than", "Equal to (not simply ‘=’)", "Greater than or equal to, binary", "Less than or equal to, binary", "Sequence (in model formulae: interaction)", "List subset, including a column in a data frame")
))
```
<!-- Is there a reason that > is here but < is not? -->

### Some notes on writing readable scripts
<iframe src="https://www.youtube-nocookie.com/embed/qqz_D1vzS5M?modestbranding=1&rel=0" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" sandbox="allow-scripts allow-same-origin" allowfullscreen="" width="500" height="281"></iframe>

Commands can all be put in the console (the one at the bottom-left of RStudio, starting with the “>”). But since you often want to redo the same thing (trying and tweaking), we urge you to store the commands in a _script_. Open a new script in RStudio (file -> new file -> R script). This will appear in the upper-left pane. Type all commands in the script. This script can be stored (extension is .R) and is basically a .txt file.

When writing a script, start with a title, preceded with the #, because it is not a command, on top, for example:

```{r eval=F}
# This script is for computing the mean maximum temperature for several days

#This vector includes the highest temperatures recorded on several days
temperature <- c(19, 17, 20, 20, 13, 13, 15, 17)
temperature %>% mean() # to compute the mean maximum temperature
```

When writing commands in the script, always put a space after a comma, and never before a comma, just like in regular English. Never put spaces inside or outside parentheses for regular function calls. Most infix operators (==, +, -, <-, etc.) should always be surrounded by spaces. The pipe operator that was used in the example above, is explained later in this document. For now, it is important to know that when indenting after a pipe `>%>` operator, you need to use two spaces (you can automatically set this in RStudio’s preferences (RStudio / Preferences / Code/ Insert spaces for tab / tab width = 2). Use `"` for quoting text. Only use `'` when the text already contains double quotes and no single quotes. If the arguments to a function don’t all fit on one line (of max 80 characters), put each argument on its own line and indent with two spaces.

### Downloading and activating R Packages
<iframe src="https://www.youtube-nocookie.com/embed/3RWb5U3X-T8?modestbranding=1&rel=0" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" sandbox="allow-scripts allow-same-origin" allowfullscreen="" width="500" height="281"></iframe>

R and RStudio can be used more efficiently, by ‘add ons’ called _packages_. These packages simplify programming in R as they include new functions that you can use. Below, you will find a list of packages to be downloaded for the various courses at BMS. After downloading (= installing) the package, there is no need to do that every time you are using R, although it may be wise to sometimes check for updates. After installing and opening R and RStudio, you can install packages using a command in either the console or in a script.

```{r eval=F}
# Installing the packages 'foreign' and 'tidyverse' (both to be introduced later)# is done with the following command in R:
install.packages("foreign", "tidyverse")
```
<!-- Changed haven to foreign, since I remember later on in the year we changed to using that for importing SPSS files -->

Downloading (= installing) a package is not the same as ‘calling’ that package for usage in R: packages are not automatically opened when you are using R and RStudio. And that is a good thing: packages can use conflicting short cuts (using functions with the same name, but with very different outcomes). Therefore, each time you are using a specific _package_ you have to load that package into the library (aka ‘activate’). A script therefore often starts with ‘calling’ the relevant packages for that script.

```{r eval=F}
# Loading the packages into the library is done with commands like these:
library(foreign)
library(tidyverse)
```

### R packages to be used at BMS
A lot of statistical analyses can be done by using ‘base R’ and its associated functions, but for some more specialized analyses we will use specialized packages.

Since the number of packages is huge[^4], we have decided to use only a limited number of packages at BMS. The use of R at BMS will be based on the _tidyverse_[^5]. We will therefore mainly use packages belonging to the _tidyverse_ (that are installed and loaded when installing and loading the _tidyverse_ package). We will also use some other packages.

[^4]:
See here all available packages: https://cran.r-project.org/web/packages/available_packages_by_name.html

[^5]:
See here the tidyverse style guide: https://style.tidyverse.org/

Here is an overview of the packages you will be using:

* `ggplot2` to visualize your data (part of the tidyverse core);
* `dplyr` and `tidyr` to change data, and to make sure that each variable is in a column, each observationis a row and each value is a cell (both part of the tidyverse core);
* `readr` to import .csv data (part of the tidyverse core);
* `readxl` to import .xls and .xlsx sheets;
* `foreign` to import SPSS, Stata, and SAS data;
* `magrittr` to provide the pipe %>% operator, used throughout the tidyverse. This will be explained later;
* `janitor` to create frequeny tables and cross tabulations (not part of the tidyverse, but aligned);
* `lme4` for the linear mixed model;
* `psych` for psychometric analysis, including scale construction and factor analysis;
* `lmerTest` to approximate _p_-values;
* `broom` to use a `tidy()` function, which may come handy when doing statistical analysis;
* `modelr` to make plots with the residuals and/or the predicted values (this package will soon be replaced by a new set of packages called ‘tidymodels’.)

At the beginning of a meeting or course we will tell you which packages to download and call.

### Datasets in R-packages
A part from a set of commands, most packages also contain data sets, to enable easy illustrations of what packages can do.

```{r eval=F}
# to see which datasets are available in the packages loaded in the library, use:
data()

# to see all 'available' data in installed packages on your computer, use:
data(package =.packages(all.available = TRUE))

# sometimes it makes sense to put a data set in the Global Environment:
gss_cat <- gss_cat # loaded in the package forcats, which is part of the core tidyverse (see below)
```

### The ‘pipe’ operator
<iframe src="https://www.youtube-nocookie.com/embed/9yjhxvu-pDg?modestbranding=1&rel=0" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" sandbox="allow-scripts allow-same-origin" allowfullscreen="" width="500" height="281"></iframe>

Some packages add additional operators to the set of base R operators. The ‘pipe’ operator is the most important one and is part of the `magrittr` package (which is part of tidyverse, but not part of the core, so it has to be put into the library separately). We urge you to use the ‘pipe’ function as much as possible.
<!-- As far as I experienced, the magrittr package doesn't have to be put in the library seperatly -->

Suppose we wanted to create a new object with some (only some) gss_cat data from 2000 after we imported these data as an object into R. In this data set all rows are observations. In this case individuals: each row is an individual in a specific year. Religion and partyid are variables in this dataset. Suppose we want to select only **data** from 2000, and focus on age and religion, and we want to store these data in a new (smaller) object called ‘gss_cat_small’. We could use:

```{r eval=F}
gss_cat <- gss_cat
gss_cat_small <- filter(select(gss_cat, year, age, relig), year==2000)
```

This is called a nested command (with the functions ‘filter’ and ‘select’). A nested command is often difficultto understand. Much simpler is:

```{r eval=F}
gss_cat_small <- gss_cat %>% 
  select(gss_cat, year, age, relig) %>% 
  filter(year==2000)
```

This command means: you have a big data frame (gss_cat, which is an object in the global environment), you select some variables from this data frame, and then you filter those cases (from a specific year).

When writing code, use the pipe `%>%` operator as much as possible. There is special hotkey in RStudio for the pipe operator: Ctrl+Shift+M (Windows & Linux), Cmd+Shift+M (Mac). We will teach you how to systematically use the pipe operator in the various units.

## Examples {.tabset}

### Example 1.R

<a href="#" class="btn btn-primary" style="float:right;"><i class="fas fa-download"></i> Download</a>

An example of code can be used including its results. The filename of the example should be descriptive of what it does. For every example, there is also a downloadable file, which also includes some extra code for the importing/installing for libraries, so the whole code can be run instantly without any modifications.

Short explanation of what the goal is of the code below

```{r eval=F}
prediction_wc <- predict(model, newdata = wc_test, type = "response")
prediction_wc <- prediction_wc > .5

tibble(predicted = prediction_wc, truth = wc_test$is_me) %>% 
tabyl(predicted, truth) %>% 
adorn_title("combined")
```
```
##  predicted/truth  0  1
##            FALSE 39 30
##             TRUE  1  4
```

### Example 2.R

<a href="#" class="btn btn-primary" style="float:right;"><i class="fas fa-download"></i> Download</a>

Another example

```{r eval=F}
corpus_a <- corpus_a %>%
  tm_map(stemDocument)

# Check first document
writeLines(as.character(corpus_a[[1]]))
```
```
## Wednesday the 28th I'm available, so I'm total okay with that
```

### Dataset.csv

<a href="#" class="btn btn-primary" style="float:right;"><i class="fas fa-download"></i> Download</a>

The (first few lines of the) dataset that is used in the examples. Preferably, the same dataset is used in all examples, but different parts of it are used.

```{r, echo=F, message=F}
library(tidyverse)
data(starwars)
starwars
```

## Functions

A short overview of the functions used in this page, which package they are in, and what they are used for, with a link to their official documentation

* [`c()`](https://www.rdocumentation.org/packages/base/topics/c)  
_package: utils (base)_   
Combine Values into a Vector or List.
* [`data()`](https://www.rdocumentation.org/packages/utils/topics/data)  
_package: utils (base)_   
Loads specified data sets, or list the available data sets.
* [`filter()`](https://www.rdocumentation.org/packages/dplyr/topics/filter)  
_package: dplyr (tidyverse)_   
Find rows where conditions are true
* [`install.packages()`](https://www.rdocumentation.org/packages/utils/topics/install.packages)   
_package: utils (base)_   
Download and install packages from the internet on your computer
* [`libray()`](https://www.rdocumentation.org/packages/base/topics/library)   
_package: base_   
Loads a package into a specific project
* [`mean()`](https://www.rdocumentation.org/packages/base/topics/mean)  
_package: utils (base)_   
Find the mean of a list of numbers.
* [`select()`](https://www.rdocumentation.org/packages/dplyr/topics/select)  
_package: dplyr (tidyverse)_   
Keeps only the variables you mention.

## FAQ
::: {#FAQ-descriptive-name}
**This is a question a student asked?**   
This is the answer to that question
:::

::: {#FAQ-another-question}
**This is another question**   
All questions have descriptive ids, so they can easily be [linked to](#FAQ-another-question) from anther page 
:::

## Resources
* [RStudio website](https://www.rstudio.com/products/rstudio/)   
More explanation about RStudio
* [Available CRAN Packages By Name](https://cran.r-project.org/web/packages/available_packages_by_name.html)   
A list of all existing packages that can be loaded into R
* [Tidyverse Style Guide](https://style.tidyverse.org/)   
Describes the style used throughout the tidyverse

## Practice
A question students should know the answer to based on the material above
<details>
  <summary>Hint 1</summary>
  
  A hint for if the students can't think of the answer themselves to get them into the right direction
</details>
<details>
  <summary>Hint 2</summary>
  
  Another hint
</details>
<details>
  <summary>Answer</summary>
  
  <a href="#" class="btn btn-primary" style="float:right;"><i class="fas fa-download"></i> Download</a>
  
  The answer with full explanation which can be downloaded
  
  ```{r eval=F}
  corpus_a <- corpus_a %>%
    tm_map(stemDocument)
  
  # Check first document
  writeLines(as.character(corpus_a[[1]]))
  ```
  ```
  ## Wednesday the 28th I'm available, so I'm total okay with that
  ```
</details>
