---
title: "Introducing R and RStudio"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This page contains information on how to install R and RStudio, what it is used for and some basic termonogoly and commands, like installing packages.

# Explanation
```{r child="book/chapter1.Rmd", echo=FALSE}
```

# Examples {.tabset}

*No examples available yet*

# Functions

A short overview of the functions used in this page, which package they are in, and what they are used for, with a link to their official documentation

* [`c()`](https://www.rdocumentation.org/packages/base/topics/c)  
_package: utils (base)_   
Combine Values into a Vector or List.
* [`data()`](https://www.rdocumentation.org/packages/utils/topics/data)  
_package: utils (base)_   
Loads specified data sets, or list the available data sets.
* [`filter()`](https://www.rdocumentation.org/packages/dplyr/topics/filter)  
_package: dplyr (tidyverse)_   
Find rows where conditions are true
* [`install.packages()`](https://www.rdocumentation.org/packages/utils/topics/install.packages)   
_package: utils (base)_   
Download and install packages from the internet on your computer
* [`libray()`](https://www.rdocumentation.org/packages/base/topics/library)   
_package: base_   
Loads a package into a specific project
* [`mean()`](https://www.rdocumentation.org/packages/base/topics/mean)  
_package: utils (base)_   
Find the mean of a list of numbers.
* [`select()`](https://www.rdocumentation.org/packages/dplyr/topics/select)  
_package: dplyr (tidyverse)_   
Get a specific column or columns from a dataset

# FAQ
*No Frequently Asked Questions yet*

# Resources
* [RStudio website](https://www.rstudio.com/products/rstudio/)   
More explanation about RStudio
* [Available CRAN Packages By Name](https://cran.r-project.org/web/packages/available_packages_by_name.html)   
A list of all existing packages that can be loaded into R
* [Tidyverse Style Guide](https://style.tidyverse.org/)   
Describes the style used throughout the tidyverse