---
title: "Handling data files"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This page will show you how you van import different types of datasets in R and how you can process its data.

# Explanation
```{r child="book/chapter2.Rmd"}
```

# Examples {.tabset}

*No examples available yet*

# Functions

A short overview of the functions used in this page, which package they are in, and what they are used for, with a link to their official documentation

* [`add_row()`](https://www.rdocumentation.org/packages/tibble/topics/add_row)  
_package: tibble (tidyverse)_   
Add a row to an existing dataset
* [`as.character()`](https://www.rdocumentation.org/packages/base/topics/logical)  
_package: base_   
Converts the data type of a variable or object into a character (text)
* [`as.factor()`](https://www.rdocumentation.org/packages/base/topics/factor)  
_package: base_   
Converts the data type of a variable or object into a factor (categories)
* [`as.integer()`](https://www.rdocumentation.org/packages/base/topics/integer)  
_package: base_   
Converts the data type of a variable or object into a integer (whole numbers)
* [`as.logical()`](https://www.rdocumentation.org/packages/base/topics/logical)  
_package: base_   
Converts the data type of a variable or object into a locical variable (TRUE or FALSE)
* [`as.numeric()`](https://www.rdocumentation.org/packages/base/topics/numeric)  
_package: base_   
Converts the data type of a variable or object into a number (real number)
* [`attr()`](https://www.rdocumentation.org/packages/base/topics/attr)  
_package: base_   
Get or set specific attributes of an object.
* [`case_when()`](https://www.rdocumentation.org/packages/dplyr/topics/case_when)  
_package: dplyr (tidyverse)_   
Return one or an other variable based on a condition
* [`colnames()`](https://www.rdocumentation.org/packages/base/topics/row+colnames)  
_package: base_   
Only get the column names of a data set
* [`data.frame()`](https://www.rdocumentation.org/packages/base/topics/data.frame)  
_package: base_   
Creates a dataframe from multiple lists
* [`filter()`](https://www.rdocumentation.org/packages/dplyr/topics/filter)  
_package: dplyr (tidyverse)_   
Get specific rows from the dataset based on certain conditions
* [`getwd()`](https://www.rdocumentation.org/packages/base/topics/getwd)  
_package: base_   
Tells you what the current working directory is
* [`is.na()`](https://www.rdocumentation.org/packages/base/topics/NA)  
_package: base_   
Check if a value is equal to `NA`
* [`read.csv()`](https://www.rdocumentation.org/packages/utils/topics/read.table)  
_package: utils (base)_   
Lets you import a .sav file made for SPSS into R as a dataset
* [`read_spss()`](https://www.rdocumentation.org/packages/haven/topics/read_spss)  
_package: haven_   
* [`rnorm()`](https://www.rdocumentation.org/packages/stats/topics/Normal)  
_package: stats (base)_   
Creates random numbers with a normal distribution
Lets you import a .sav file made for SPSS into R as a dataset
* [`sd()`](https://www.rdocumentation.org/packages/stats/topics/sd)  
_package: stats_   
Find the standard deviation of a list of numbers.
* [`select()`](https://www.rdocumentation.org/packages/dplyr/topics/select)  
_package: dplyr (tidyverse)_   
Get a specific column or columns from a dataset
* [`setwd()`](https://www.rdocumentation.org/packages/base/topics/getwd)  
_package: base_   
Sets the working directory (the folder in which R will look for your files)
* [`str()`](https://www.rdocumentation.org/packages/utils/topics/str)  
_package: utils (base)_   
Get a quick overview of the types of data in your matrix and their names
* [`sum()`](https://www.rdocumentation.org/packages/base/topics/sum)  
_package: base_   
Find the sum of a list of numbers.
* [`summary()`](https://www.rdocumentation.org/packages/base/topics/summary)  
_package: base_   
Give a summary of all variables in a dataset, giving information like minimum, maximum, mean, and frequencies
* [`left_join()`](https://www.rdocumentation.org/packages/stats/dplyr/mutate-joins)  
_package: dplyr (tidyverse)_   
Merge two datasets together using a variable they have in common
* [`mean()`](https://www.rdocumentation.org/packages/base/topics/mean)  
_package: utils (base)_   
Find the mean of a list of numbers.
* [`mutate()`](https://www.rdocumentation.org/packages/dplyr/topics/mutate)  
_package: dplyr (tidyverse)_   
Change the values of an entire column in a dataset
* [`na_if()`](https://www.rdocumentation.org/packages/dplyr/topics/na_if)  
_package: dplyr (tidyverse)_   
Change a value to `NA` if it is equal to a specific value
* [`pivot_longer()`](https://www.rdocumentation.org/packages/tidyr/topics/pivot_longer)  
_package: tidyr (tidyverse)_   
Change a dataset from short format to long format
* [`pivot_wider()`](https://www.rdocumentation.org/packages/tidyr/topics/pivot_wider)  
_package: tidyr (tidyverse)_   
Change a dataset from long format to short format
* [`rename()`](https://www.rdocumentation.org/packages/dplyr/topics/sd)  
_package: dplyr (tidyverse)_   
Changes the name of a variable in a dataset
* [`rowSums()`](https://www.rdocumentation.org/packages/base/topics/colSums)  
_package: base_   
Get the sum of the values of the rows in a dataset
* [`View()`](https://www.rdocumentation.org/packages/utils/topics/View)  
_package: utils (base)_   
Shows a dataframe as a table in RStudio

# FAQ
::: {#faq-file-not-found}
**When trying to open a dataset I get the message `Error : no such file or directory`**  
When loading a file, R will only search for that file in one directory: The working directory. You can find out what the working directory is by using the command `getwd()`. If that is not the folder where your file is in, you need to change it. You can do that using `setwd()`. When typing the path of this directory, you can't use the default backslash (\\) that is normally used in paths. Therefore, you need to change it to either a forward slash (/) or a double backslash (\\\\). So if your file is in the folder "C:\\Users\\username\\Downloads", you type `setwd("C:\\Users\\username\\Downloads")`. If you do not know the full path of the folder, you can type `rstudioapi::selectDirectory()` in your console. This will allow you to pick a folder by hand, and will give you the full path you need to use in your command with the slashes already changed.

If the file still cannot be found, check for any typo's in the folder path or filename. When typing the filename, make sure to include the file extension. This is for example `.csv` or `.sav`. File extensions may be hidden in your file explorer, making them difficult to know. To see file extensions on Windows, open file explorer, go to the *View* tab, and under *Show/Hide* check the box *File name extensions*. On Mac, select a file, then choose *File > Get Info*. Click the arrow next to *Name & Extension* and deselect *Hide extension*
:::
::: {#faq-open-sav}
**My computer can’t open .sav files. I don’t have the right program installed.**  
You don’t need to open .sav files directly with your computer. You only have to import the sav files into R. You can do that using the `foreign` library and the `read.spss()` command. 
:::

# Resources
* [Data transformation cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/master/data-transformation.pdf)  
An overview of all the functions you can use to find and manipulate data in your dataframe.