---
title: "Univariate data analysis and data visualization"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This page will show you how you can analyze and visualize single variables using R.

# Explanation
```{r child="book/chapter3.Rmd"}
```

# Examples {.tabset}
*No examples available yet*

# Functions
* [`adorn_pct_formatting()`](https://www.rdocumentation.org/packages/janitor/topics/adorn_totals)  
_package: janitor_   
Change numeric columns to percentages in a frequency table (tabyl).
* [`adorn_totals()`](https://www.rdocumentation.org/packages/janitor/topics/adorn_totals)  
_package: janitor_   
Add a row or column with totals to a frequency table (tabyl).
* [`aes()`](https://www.rdocumentation.org/packages/ggplot2/topics/aes)  
_package: ggplot2 (tidyverse)_   
Create an aestetic map, which explains which visual properties (aesthetics) are linked (mapped) to which variables in the dataframe.
* [`ggplot()`](https://www.rdocumentation.org/packages/ggplot2/topics/ggplot)  
_package: ggplot2 (tidyverse)_   
Initialize any type of graph.
* [`geom_bar()`](https://www.rdocumentation.org/packages/ggplot2/topics/geom_bar)  
_package: ggplot2 (tidyverse)_   
Create a bar chart using the initialized ggplot.
* [`geom_boxplot()`](https://www.rdocumentation.org/packages/ggplot2/topics/geom_boxplot)  
_package: ggplot2 (tidyverse)_   
Create a boxplot using the initialized ggplot.
* [`geom_histogram()`](https://www.rdocumentation.org/packages/ggplot2/topics/geom_histogram)  
_package: ggplot2 (tidyverse)_   
Create a histogram using the initialized ggplot.
* [`map()`](https://www.rdocumentation.org/packages/purrr/topics/map)  
_package: purrr (tidyverse)_   
Apply a transformation to every variable in the dataframe.
* [`max()`](https://www.rdocumentation.org/packages/base/topics/max)  
_package: base_   
Find the highest number from a list of numbers.
* [`mean()`](https://www.rdocumentation.org/packages/base/topics/mean)  
_package: utils (base)_   
Find the mean of a list of numbers.
* [`median()`](https://www.rdocumentation.org/packages/stats/topics/median)  
_package: stats (base)_   
Find the median of a list of numbers.
* [`min()`](https://www.rdocumentation.org/packages/base/topics/min)  
_package: base_   
Find the lowest number from a list of numbers.
* [`sd()`](https://www.rdocumentation.org/packages/stats/topics/sd)  
_package: stats (base)_   
Find the standard deviation of a list of numbers.
* [`summarise()`](https://www.rdocumentation.org/packages/dplyr/topics/summarise)  
_package: dplyr (tidyverse)_   
Create a new dataframe summarizing the values in another dataframe.
* [`tabyl()`](https://www.rdocumentation.org/packages/janitor/topics/tabyl)  
_package: janitor_   
Generate a frequency table.
* [`var()`](https://www.rdocumentation.org/packages/stats/topics/var)  
_package: stats (base)_   
Find the variation of a list of numbers.

# FAQ
::: {#faq-pipe-for-plus}
*When creating a graph, i get the message `Did you use %>% instead of +?`*
When using the `ggplot()` command together with another command like `geom_bar()` or `geom_boxplot()` you don't use the pipe (`%>%`) between them, but instead you use the `+`. That is because these two commands need to be executed together, and instead of that the result of one needs to be used in the next.
:::

# Resources
* [Data visualisation cheatsheet with ggplot2](https://raw.githubusercontent.com/rstudio/cheatsheets/master/data-visualization.pdf)  
An overview of all the different graph types and how to make them using the `ggplot()` function.