---
title: "Template Page"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
A short explanation on what the information on this page will help you with, so people can see if this page contains the information they are looking for

## Explanation

Explanation about how to do the thing, using both videos and text. In the text functions and other parts of codes are reffered to like_this(). For bigger parts of code, that will look
``` code
Like this
```
```
And code results look like this
```

To get the code to look nice, you can add it to a RMarkdown HTML document, and copy the HTML directly from there

<iframe src="https://www.youtube-nocookie.com/embed/videoseries?playlist=au6IwOKJ4Yo,KErEm6rmbBA&amp;modestbranding=1&amp;rel=0" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" sandbox="allow-scripts allow-same-origin" allowfullscreen="" width="500" height="281"></iframe>

When multiple videos belong together, these are put in a playlist, rather then shown next to each other. For the embed the domain www.youtube-nocookie.com is used instead of the normal www.youtube.com domain, because everyone hates it when youtube suddenly recommends them videos about statistics after they've watched videos for uni. A sandbox is added and modest branding and related videos are blocked, to avoid people just ending up browsing YT.

### Subheadings

For longer explanations, it is advisable to use subheadings, with descriptive id's so they can easily be [linked to](#subheadings)

## Examples {.tabset}

### Example 1.R

<a href="#" class="btn btn-primary" style="float:right;"><i class="fas fa-download"></i> Download</a>

An example of code can be used including its results. The filename of the example should be descriptive of what it does. For every example, there is also a downloadable file, which also includes some extra code for the importing/installing for libraries, so the whole code can be run instantly without any modifications.

Short explanation of what the goal is of the code below

```{r eval=F}
prediction_wc <- predict(model, newdata = wc_test, type = "response")
prediction_wc <- prediction_wc > .5

tibble(predicted = prediction_wc, truth = wc_test$is_me) %>% 
tabyl(predicted, truth) %>% 
adorn_title("combined")
```
```
##  predicted/truth  0  1
##            FALSE 39 30
##             TRUE  1  4
```

### Example 2.R

<a href="#" class="btn btn-primary" style="float:right;"><i class="fas fa-download"></i> Download</a>

Another example

```{r eval=F}
corpus_a <- corpus_a %>%
  tm_map(stemDocument)

# Check first document
writeLines(as.character(corpus_a[[1]]))
```
```
## Wednesday the 28th I'm available, so I'm total okay with that
```

### Dataset.csv

<a href="#" class="btn btn-primary" style="float:right;"><i class="fas fa-download"></i> Download</a>

The (first few lines of the) dataset that is used in the examples. Preferably, the same dataset is used in all examples, but different parts of it are used.

```{r, echo=F, message=F}
library(tidyverse)
data(starwars)
starwars
```

## Packages

A short overview of the packages used in this page, and what they are used for, with a link to their official documentation

* [Package](https://www.rdocumentation.org/packages/tidyverse/)   
What the Package in question exactly does
* [Package](https://www.rdocumentation.org/packages/tidyverse/)   
What the Package in question exactly does
* [Package](https://www.rdocumentation.org/packages/tidyverse/)   
What the Package in question exactly does

## Functions

A short overview of the functions used in this page, and what they are used for, with a link to their official documentation

* [Function](https://www.rdocumentation.org/packages/ggplot2/versions/3.3.3/topics/ggplot)   
What the Function in question exactly does
* [Function](https://www.rdocumentation.org/packages/ggplot2/versions/3.3.3/topics/ggplot)   
What the Function in question exactly does
* [Function](https://www.rdocumentation.org/packages/ggplot2/versions/3.3.3/topics/ggplot)   
What the Function in question exactly does

## FAQ
::: {#FAQ-descriptive-name}
**This is a question a student asked?**   
This is the answer to that question
:::

::: {#FAQ-another-question}
**This is another question**   
All questions have descriptive ids, so they can easily be [linked to](#FAQ-another-question) from anther page 
:::

## Resources
* [Name of resource](#)   
Explanation of what you can find on said resource
* [Name of resource](#)   
Explanation of what you can find on said resource
* [Name of resource](#)   
Explanation of what you can find on said resource

## Practice
A question students should know the answer to based on the material above
<details>
  <summary>Hint 1</summary>
  
  A hint for if the students can't think of the answer themselves to get them into the right direction
</details>
<details>
  <summary>Hint 2</summary>
  
  Another hint
</details>
<details>
  <summary>Answer</summary>
  
  <a href="#" class="btn btn-primary" style="float:right;"><i class="fas fa-download"></i> Download</a>
  
  The answer with full explanation which can be downloaded
  
  ```{r eval=F}
  corpus_a <- corpus_a %>%
    tm_map(stemDocument)
  
  # Check first document
  writeLines(as.character(corpus_a[[1]]))
  ```
  ```
  ## Wednesday the 28th I'm available, so I'm total okay with that
  ```
</details>
