---
title: "Univariate data analysis and data visualization"
output: html_fragment
---

A variable can be described with statistics like the mode, mean, median, variance and standard deviation. Also, it is interesting to visualize a variable using a plot.

## Summarizing numerical variables: mean, median, variation and standard deviation

Below you will see several examples of obtaining descriptive statistics for items in a data set called `data_new`.

```{r eval = F, tidy = TRUE}

# To summarise some main characteristics of one numerical item (called Item1):
data_new %>% 
  summarise(mean = mean(Item1),
            sd = sd(Item1),
            var = var(Item1),
            minimum = min(Item1),
            maximum = max(Item1)
            )
# You can also ask for only one of the statistics.

# To summarize the main characteristics (mean, median, minimum, maximum, Q1, Q3 and number of missings) of multiple variables (in this example: the first 10 variables) in the data set:
data_new %>% 
  select(1:10) %>%
  summary()           

# To compute a certain statistic for the first 10 variables in the data set, you can also use the map() function:

data_new %>%
  select(1:10) %>%
  map(var)

# instead of var, you can also use mean, sd, or other functions

```


## Frequency tables with janitor: tabyl

The tabyl() function from the `janitor` package is a way to get frequency tables. tabyl() is tidyverse-aligned and is primarily built upon the dplyr and tidyr packages.

```{r eval = F, tidy = TRUE}

# Make sure 'janitor' in loaded into the library

# Create a frequency table
gss_cat %>%
  tabyl(race)

# Create frequency tables for all variables of a data set:
gss_cat %>%
  map(tabyl)

# If the values are in alphabetical, rather than in a meaningful order, make sure the variable is stored as an 'ordered factor' (see above).

# The frequency table can be made nicer using the adorn commands, for example:
gss_cat %>%
  tabyl(marital) %>%
  adorn_totals("row") %>%
  adorn_pct_formatting()

```

## Univariate graphs: creating bar charts, box plots and histograms using ggplot

We will use the package `ggplot2` (part of the core tidyverse) for creating visualizations. The function **ggplot** is extremely flexible and there is an almost infinite number of ways to display your data.

The basic idea of ggplot commands is that you 
(1) have a data frame, pipe (%>%) that into a 
(2) ggplot(), add a
(3) geom_...() to select the type of display you want to have, and
(4) use aesthetics (aes()) to select the variable(s) you will use from that data frame.

There are more 'layers' in a ggplot, but these are the most important.

For example, you can use a bar chart to visualize a categorical (= nominal) variable. 

```{r, creating barcharts, boxplots and histograms, eval = F}

# creating a basic bar chart with the nominal variable item 1
gss_cat %>% 
  ggplot() +
  geom_bar(aes(x = marital))

# to create a box plot for the ratio variable tvhours in gss_cat 
# (please note that x = and y = in the aesthetics tilt the picture)

gss_cat %>% 
  ggplot() +
  geom_boxplot(aes(y = tvhours))

# to create a histogram for the ratio variable tvhours in gss_cat
gss_cat %>% 
  ggplot() +
  geom_histogram(aes(x = tvhours))

# there are many more geom's to use, but we will focus on these in univariate analysis. But this is definitely a topic you can explore further. Think for example about changing the colors of the bars in a barchart, adding better titles or adding annotations.

# Note 1: although we generally recommend using the pipe operator like this, if you start visualizing data from different objects in one plot, it may be wise to put the data inside the ggplot:

ggplot(gss_cat) +
  geom_histogram(aes(x = tvhours))

# Note 2: the aesthetics are put in the geom. However, if you combine different geom's (scatter and line) displaying the same variables, it is more efficient, to but the aesthetics in the general ggplot command too:

ggplot(gss_cat, aes(x = tvhours)) +
  geom_histogram()
  
```
  
## Univariate inferential statistics

Confidence intervals and tests for one variable can be constructed for dichotomies (proportions), nominal variables (goodness of fit) and interval/ratio variables (means testing).

```{r, confidence intervals and tests for one variable, eval = F}

# For a proportion, use
n_total <- 1000 # total number of (valid) cases
n_pos <- 690 # number of positives
prop.test(n_pos, n_total, alternative = "two.sided", conf.level = 0.95) 
# for confidence interval and two sided test

# For a goodness of fit test, use
n_observed <- c(166, 142, 80, 160)
p_expected <- c(0.25, 0.25, 0.25, 0.25) # these have to add to 1 (proportions)
g_o_f <- chisq.test(n_observed, p = p_expected)
g_o_f

# For a test of the median, use
# assuming the data contains a vector with a difference, called data$t2_t1
wilcox.test(data$t2_t1, alternative = "two.sided")
# this can be done using the idea of a signed rank and the linear model too.
# assuming you constructed the signed ranks yourself, use ...
data %>% lm(signed_rank_t2_t1 ~ 1, .)

# For means testing use either t.test or the lm command
# assuming the data are stored in a data frame as data$var
mu_1 <- 100 # if you want to check whether it is different from 100

t.test(data$var, mu = mu_1)

data %>% 
  lm((var - mu_1) ~ 1, .) %>% 
  summary

# For finding the p-value for a specific t-value
# instead of using a table, use R:
q <- 1.96 # the t-value
df <- 499 # the degrees of freedom
pt(q, df, lower.tail = FALSE)
# NOTE: this gives the percentage on ONE side
```
