---
title: "Intro R for BMS Bachelor programs"
author: "OMD staff"
date: "version 25/05/2020"
output:
  pdf_document
---

\newpage
\tableofcontents
\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80),echo = TRUE, fig.align = "left")
library(tidyverse)
library(formatR) #in combination with tidy = TRUE in specific codes chuncks this wraps the text.

```

\newpage
# Introduction

This document describes how bachelor students in BMS programs will use the program language R in the courses for Research Methods and Statistics (and beyond). The document will be available at all time when making assignments and when making exams in which R is used. The document will be updated when the staff discovers problems in the way students deal with R (check the version date on top of this document).


## How to use this document?

There is no need to learn this document 'by heart'. It is mainly a reference guide. Read until "Installing R and RStudio on your laptop" once before doing anything else with R. Then install R and Rstudio. After that read the rest of "1. Introducing R and RStudio" once.
After this make some of the assignments we created. Use the materials in this reference guide, in case you get stuck. After doing three of four R assignments, read the first chapter again, and skim through the remainder of the document.
Before you start using R in other courses, read through the document again, to re-familiarize yourself with R.

\newpage

# 1. Introducing R and RStudio
```{r child="chapter1.Rmd", echo=FALSE}
```

\newpage
# 2. Handling data files
```{r child="chapter2.Rmd", echo=FALSE}
```


\newpage
# 3. Univariate data analysis and data visualization
```{r child="chapter3.Rmd", echo=FALSE}
```

\newpage
# 4. Bivariate data analysis and bivariate data vizualisation
```{r child="chapter4.Rmd", echo=FALSE}
```

\newpage
# 5. Multivariate data analysis
```{r child="chapter5.Rmd", echo=FALSE}
```

\newpage
# 6. Psychometric analyses
<<<<<<< HEAD
```{r child="chapter6.Rmd", echo=FALSE}
=======
```{r child="chapter5.Rmd", echo=FALSE}
>>>>>>> 83c5080d8e3ae2a91818a3d631736c843af2ef9d
```